Modal Abstractions for Virtualizing Memory Addresses

### 2024-08-05 examples.tex
working into 5.1.5, draft fix to first line of fig 11

;;pte_t *pte_get_next_table_succ(pte_t *entry) {

@@

jne    161 <pte_get_next_table_succ+0xa1> (* \label{line:check_entry_present_jump} *) ;; Jump if the present bit is not zero

@@

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
... ;; Continued from Figure (*\ref{fig:calltopteinitialize}*) either by following the pte allocation path or not
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \}),m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{rcx}  \mapsto_{\textsf{r}} \textsf{\_}  }_{\rtv}$
$\specline{ \textsf{entry} \mapsto_{\textsf{id}} \textsf{\_} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{(pte\_initialized (entry\_val.pfn))} \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{(pte\_initialized (entry\_val.pfn)))} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{ table\_root (pte\_initialize(entry\_val.pfn))} }_{\rtv}$
;;uintptr_t next_virt_addr = (uintptr_t) P2V(entry.pfn<<12); (*\label{line:p2v}*) 
 movabs KERNBASE,rcx
 add    rcx,rax
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \}),m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{rcx}  \mapsto_{\textsf{r}} \textsf{KERNBASE} }_{\rtv}$
$\specline{ \textsf{entry} \mapsto_{\textsf{id}} \textsf{\_} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{(pte\_initialized (entry\_val.pfn))} \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{(pte\_initialized (entry\_val.pfn)))} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{ table\_root (pte\_initialize(entry\_val.pfn)) + KERNBASE} }_{\rtv}$
...
;;next = (pte_t *) next_virt_addr;
;;clean up the stack and return next
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates). Abreviating the other relevant resources (including the lower entries) peeled-out of the invariant as $\textsf{R}_{\textsf{children}}$ }
\label{fig:p2v}
\end{figure}
Figure \ref{fig:p2v} shows the verification of the end of \lstinline|pte_get_next_table| specialized to the case where 
where no allocation was necessary (i.e., the conditional on Line \ref{line:check_entry_present} of Figure \ref{fig:calltopteinitialize} was taken).
The code loads \lstinline|rcx| with the offset value \textsf{KERNBASE}, which gives us the value of the virtual address ($\textsf{entry}_{\textsf{pfn}}$ \textsf{+KERNBASE})
of the \emph{base} of the next level of the page table.
\todo[inline]{the next sentence depends on having figure 10 updated to reflect the page-worth of tokens}
While we could now convert this address to a virtual points-to, this is not necessarily the correct thing to do.
The caller \lstinline|walkpgdir| (discussed next) uses \lstinline|pte_get_next_table| to retrieve just the base address,
because only the caller knows which entry in the subsequent table will be accessed (it depends on the corresponding bits from the virtual
address being translated). So instead we pass back the per-address-space invariant with the identity mapping resources for \lstinline|entry|
pulled out. It is up to the caller to determine which entry in that table must actually
be accessed --- by selecting the appropriate index into the 512 ghost map tokens returned in the postcondition,
and using the ghost translation and physical location portions of the invariant to assemble a vpte-pointsto.
% in the identity map ($\Xi\setminus\{entry\}$) of the kernel invariant.
% the logical update in Specification  Lines 5-10 to 10-14 for obtaining virtual-pointsto resource for the frame 
% ($\textsf{entry}_{\textsf{pfn}}$) by removing it from the ghost map ($\Xi\setminus\{entry\}\cup \{\textsf{entry}_{\textsf{pfn}}) \}$) 
% in Line 5 and compute the identity mapping for this physical frame address in Line 13 in Figure \ref{fig:p2v}).

\subsubsection{Walking Page-Table Tree: Calling \textsf{pte\_get\_nexttable} for Each Level}
\label{wlkpgdir}

\todo[inline,color=blue]{TAKEN FROM get-next-table DISCUSSION:
In the very first precondition on the client side -- e.g.\textsf{pte\_walkpgdir} in Figure \ref{walkpgdir} -- 
we see that \textsf{vpte-ptsto} for the entry being accessed (\textsf{entry}) is constructed -- in Specification Line \ref{line:get_next_vpte_precondition} of Figure \ref{fig:calltopteinitialize} for \textsf{entry}. Intuitively speaking, we start traversal from a certain entry address at the current level for which we know ($\textsf{entry} \mapsto_{\textsf{id}} \textsf{\_}$ in Line 3 in Figure \ref{walkpgdir}) that there is a memory mapping ($\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$  Line 15 in Figure \ref{walkpgdir}) that can be used to explore deeper in the page-tree.

The most crucial aspect of page-table entry access relies on the subtle interplay between updating the shared page tables, and the not violating the memory mappings. 
Concretely speaking, accessing to the entry -- whose address is kept in the register $\texttt{r8}$ -- in Line 15 in Figure \ref{fig:calltopteinitialize} requires a 
memory mapping that is logically provided through the update from Specfication Lines 3 to 15 in Figure \ref{walkpgdir} on the client side by the removal of \textsf{entry}
from the map $\Xi$ in the invariant to construct the \textsf{vpte-ptsto} (Specfication Line 15 in Figure \ref{walkpgdir}). 
This magical abstraction -- represented as $\Xi$ in the invariant definition -- provides us to construct this \textsf{pte-ptsto} out of an physical address is called \textit{identity-mappings}.
}

\todo[inline,color=blue]{MORE TO INTEGRATE: 
\textsf{pte\_walkpgdir}, as a client, holds the knowledge that there exists an identity mapping for the physical entry address (\textsf{entry}) in the root page table ($\textsf{L}_{4}$):  $\mathsf{entry} \mapsto_{\textsf{id}} \textsf{\_}$ in Specification Line 3 is a partially owned token for accessing and looking up the resources in the identity map, $\Xi$, to construct the \textit{virtual-to-physical} pointsto relation $\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry \entry\_val}$ with the virtual address (\textsf{entry+KERNBASE}) obtained by offsetting the physical address (\textsf{entry}). With this knowledge on the root-page-table-entry, we can start traversing the page-table tree which requires locating the address of the next table -- a call to \textsf{pte\_get\_next\_table} shown in Figure \ref{fig:calltopteinitialize}. 
Beyond a frame, the precondition before Line 15 requires the current address space invariant, and knowledge that \textsf{entry} is mapped to a random entry value, subtly, the operation also, at least, requires that the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
}

Realizing page-table directory walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdir}. The special part of the specficication for a page table walk can be considered as accumulation of memory mappings for the page-table entries visited and frame addresses for page-tables. For example, Line 23-24 in Figure \ref{walkpgdir} shows the virtual-pointsto for a L3 page table and the virtual-page-table-entry-pointsto for an L4 entry. In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk\_succ}}$ -- which allows us to construct the path to the L1 entry in the tree to insert a new page.  

@@

;;pte_t *walkpgdir_succ(pte_t *pml4, const void *va) {

@@

@@

;;pte_t *pdp = pte_get_next_table_succ(pml4_entry);

@@

callq  c0 <pte_get_next_table_succ>

@@

$\specline{ \textsf{plm4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4e\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv}$
$\specline{ \textsf{pdp\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3e\_val} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{pdp\_entry})}{\textsf{3}}}_{\rtv}$

@@

;;pte_t *pd = pte_get_next_table_succ(pdp_entry);

@@

;;pte_t *pt = pte_get_next_table_succ(pd_entry);
$\specline{\textsf{R}_{\textsf{walk\_acc}}}_{\rtv}$

@@

%\begin{figure}\footnotesize
%\begin{lstlisting}[mathescape]
%;;if (!entry->present) {
%;;}
%...
%$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$ 
%mov    rax,-0x18[rbp]
%;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
%mov    -0x18[rbp],rax
%movabs KERN_BASE,rcx
%add    rcx,rax
%mov    rax,-0x20[rbp]
%\end{lstlisting}
%\vspace{-1em}
%\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
%associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
%\label{fig:p2v}
%\end{figure}

@@

;;pte = walkpgdir_succ(pml4, a);

@@

callq  291 <vaspace_mappage+0x41>	28d: R_X86_64_PLT32	walkpgdir_succ-0x4

>>>>

;;pte_t *pte_get_next_table(pte_t *entry) {

@@

jne    161 <pte_get_next_table+0xa1> (* \label{line:check_entry_present_jump} *) ;; Jump if the present bit is not zero

@@

@@

... ;; Continued from Figure (*\ref{fig:calltopteinitialize}*); assertions below specialized to non-allocating path for clarity
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \}),m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{rcx}  \mapsto_{\textsf{r}} \textsf{\_}  }_{\rtv}$
$\specline{ \textsf{entry} \mapsto_{\textsf{id}} \textsf{\_} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{(pte\_initialized (entry\_val.pfn))} \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{(pte\_initialized (entry\_val.pfn)))} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{ table\_root (pte\_initialize(entry\_val.pfn))} }_{\rtv}$
$\specline{\forall_{i\in \textsf{0 ... 511} } \ldotp  \ghostmaptoken{\textsf{id}}{((\textsf{table\_root (pte\_initialized (entry\_val.pfn)))}) + \textsf{i * 8})}{\textsf{v-1}}  }$
;;uintptr_t next_virt_addr = (uintptr_t) P2V(entry.pfn<<12); (*\label{line:p2v}*) 
 movabs KERNBASE,rcx
 add    rcx,rax
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \}),m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{rcx}  \mapsto_{\textsf{r}} \textsf{KERNBASE} }_{\rtv}$
$\specline{ \textsf{entry} \mapsto_{\textsf{id}} \textsf{\_} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{(pte\_initialized (entry\_val.pfn))} \urcorner }_{\rtv}$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{(pte\_initialized (entry\_val.pfn)))} \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{ table\_root (pte\_initialize(entry\_val.pfn)) + KERNBASE} }_{\rtv}$
$\specline{\forall_{i\in \textsf{0 ... 511} } \ldotp  \ghostmaptoken{\textsf{id}}{((\textsf{table\_root (pte\_initialized (entry\_val.pfn)))}) + \textsf{i * 8})}{\textsf{v-1}}  }$

@@

;;next = (pte_t *) next_virt_addr;
;;clean up the stack and return next
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates).
 % Abreviating the other relevant resources (including the lower entries) peeled-out of the invariant as $\textsf{R}_{\textsf{children}}$
}
\label{fig:p2v}
Figure \ref{fig:p2v} shows the verification of the end of \lstinline|pte_get_next_table| specialized to the case where 
where no allocation was necessary (i.e., the conditional on Line \ref{line:check_entry_present} of Figure \ref{fig:calltopteinitialize} was taken).
The code loads \lstinline|rcx| with the offset value \textsf{KERNBASE}, which gives us the value of the virtual address ($\textsf{entry}_{\textsf{pfn}}$ \textsf{+KERNBASE})
of the \emph{base} of the next level of the page table.
% \todo[inline]{the next sentence depends on having figure 10 updated to reflect the page-worth of tokens}
While we could now convert this address to a virtual points-to, this is not necessarily the correct thing to do.
The caller \lstinline|walkpgdir| (discussed next) uses \lstinline|pte_get_next_table| to retrieve just the base address,
because only the caller knows which entry in the subsequent table will be accessed (it depends on the corresponding bits from the virtual
address being translated). So instead we pass back the per-address-space invariant with the identity mapping resources for \lstinline|entry|
pulled out. It is up to the caller to determine which entry in that table must actually
be accessed --- by selecting the appropriate index into the 512 ghost map tokens returned in the postcondition,
and using the ghost translation and physical location portions of the invariant to assemble a vpte-pointsto.
% in the identity map ($\Xi\setminus\{entry\}$) of the kernel invariant.
% the logical update in Specification  Lines 5-10 to 10-14 for obtaining virtual-pointsto resource for the frame 
% ($\textsf{entry}_{\textsf{pfn}}$) by removing it from the ghost map ($\Xi\setminus\{entry\}\cup \{\textsf{entry}_{\textsf{pfn}}) \}$) 
% in Line 5 and compute the identity mapping for this physical frame address in Line 13 in Figure \ref{fig:p2v}).

\subsubsection{Walking Page-Table Tree: Calling \textsf{pte\_get\_next\_table} for Each Level}
\label{wlkpgdir}
;;pte_t *walkpgdir(pte_t *pml4, const void *va) {
 $\specline{\textsf{P} \ast \ulcorner\textsf{rtv+KERNBASE}=\textsf{pml4}\urcorner \ast \forall_{i\in\textsf{0..511}}\ghostmaptoken{\textsf{id}}{(\textsf{rtv}+i*8)}{\textsf{4}} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}}_{\rtv}$

@@

;;pte_t *pdp = pte_get_next_table(pml4_entry);

@@

callq  c0 <pte_get_next_table>

@@

$\specline{ \textsf{plm4\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l4e\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv}$(*\label{line:ex_l4_vpte}*)
$\specline{ \textsf{pdp\_entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{l3e\_val} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{pdp\_entry})}{\textsf{3}}}_{\rtv}$(*\label{line:ex_l3_vpte}*)

@@

;;pte_t *pd = pte_get_next_table(pdp_entry);

@@

;;pte_t *pt = pte_get_next_table(pd_entry);
$\specline{\textsf{R}_{\textsf{walk}}}_{\rtv}$

@@

Realizing a software page-table walk amounts to calling \textsf{pte\_get\_next\_table} for each level as shown in Figure \ref{walkpgdir}. 
The special part of the specficication for a page table walk can be considered as accumulation of memory mappings for the page-table entries 
visited and frame addresses for page-tables. 
For example, Lines \ref{line:ex_l4_vpte} and \ref{line:ex_l3_vpte} in Figure \ref{walkpgdir} show the virtual pte-pointsto assertions for L4 and L3 entries.
In the final post-condition, we expect the accumulation of these resources from each level -- $\textsf{R}_{\textsf{walk}}$ -- 
which allows us to construct and return the path to the L1 entry in the tree to insert a new page.

This is the code which performs the actual physical-to-virtual conversions using the identity mapping portion of the per-address-space invariant.

\todo[inline]{colin frontier, need to incorporate these next 3 (2+1) paragraphs taken from earlier}
\todo[inline,color=green]{need to finish updating ghost token assertions in figure}

In the very first precondition on the client side -- e.g.\textsf{pte\_walkpgdir} in Figure \ref{walkpgdir} -- 
we see that \textsf{vpte-ptsto} for the entry being accessed (\textsf{entry}) is constructed -- in S
pecification Line \ref{line:get_next_vpte_precondition} of Figure \ref{fig:calltopteinitialize} for \textsf{entry}. Intuitively speaking, we start traversal from a certain entry address at the current level for which we know ($\textsf{entry} \mapsto_{\textsf{id}} \textsf{\_}$ in Line 3 in Figure \ref{walkpgdir}) that there is a memory mapping ($\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$  Line 15 in Figure \ref{walkpgdir}) that can be used to explore deeper in the page-tree.

The most crucial aspect of page-table entry access relies on the subtle interplay between updating the shared page tables, and the not violating the memory mappings. 
Concretely speaking, accessing to the entry -- whose address is kept in the register $\texttt{r8}$ -- in Line 15 in Figure \ref{fig:calltopteinitialize} requires a 
memory mapping that is logically provided through the update from Specfication Lines 3 to 15 in Figure \ref{walkpgdir} on the client side by the removal of \textsf{entry}
from the map $\Xi$ in the invariant to construct the \textsf{vpte-ptsto} (Specfication Line 15 in Figure \ref{walkpgdir}). 
This magical abstraction -- represented as $\Xi$ in the invariant definition -- provides us to construct this \textsf{pte-ptsto} out of an physical address is called \textit{identity-mappings}.

\textsf{pte\_walkpgdir}, as a client, holds the knowledge that there exists an identity mapping for the physical entry address (\textsf{entry}) in the root page table ($\textsf{L}_{4}$):  $\mathsf{entry} \mapsto_{\textsf{id}} \textsf{\_}$ in Specification Line 3 is a partially owned token for accessing and looking up the resources in the identity map, $\Xi$, to construct the \textit{virtual-to-physical} pointsto relation $\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry \entry\_val}$ with the virtual address (\textsf{entry+KERNBASE}) obtained by offsetting the physical address (\textsf{entry}). With this knowledge on the root-page-table-entry, we can start traversing the page-table tree which requires locating the address of the next table -- a call to \textsf{pte\_get\_next\_table} shown in Figure \ref{fig:calltopteinitialize}. 
Beyond a frame, the precondition before Line 15 requires the current address space invariant, and knowledge that \textsf{entry} is mapped to a random entry value, subtly, the operation also, at least, requires that the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

%\begin{figure}\footnotesize
%\begin{lstlisting}[mathescape]
%;;if (!entry->present) {
%  ...
%
%;;}
%...
%$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv}$ 
%mov    rax,-0x18[rbp]
 % 
%;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
%mov    -0x18[rbp],rax
%movabs KERN_BASE,rcx
%add    rcx,rax
%mov    rax,-0x20[rbp]
%\end{lstlisting}
%\vspace{-1em}
%\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
%associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
%\label{fig:p2v}
%\end{figure}

@@

;;pte = walkpgdir(pml4, a);

@@

callq  291 <vaspace_mappage+0x41>	28d: R_X86_64_PLT32	walkpgdir-0x4
